A Sound and Complete Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2025-08-19 paper/COLA/paper.tex
Fix a few typos

\author{Miguel Nunes}
\ead{miguel.alfredo.nunes@gmail.com}
\address{Santa Catarina State University}

@@

Such wide range of highly expressive logical languages is the ideal tool to
formalize the many non-trivial properties of modern advanced software systems,
such as those found in cars, airplanes, medical devices and banking applications.
By combining said modal logics with interactive proof assistants,
such as Coq~\cite{manual_coq} and Isabelle/HOL~\cite{manual_isabelle}, both of
which have been applied with success in non-trivial projects like a certified compiler
for C programming language~\cite{Leroy09} and the kernel of an operating
system~\cite{Klein10},
% it is possible to formalize and verify highly complex computer systems.
we argue we may achieve betters ways of formally describing and verifying complex programs.

@@

We define next a proof of the previous propositional logical formula
that, in contrast to the previous proof, that was built using tactics, is coded
directly as a function:
\begin{minted}{coq}
Definition example: (A -> B) -> (B -> C) -> A -> C :=
  fun (H: A -> B) (H': B -> C) (HA: A) => H' (H HA).
\end{minted}
However, even for very simple theorems, coding a definition directly
as a Coq term can be a hard task. Because of this, the use of tactics
has become the standard way of proving theorems in Coq. Furthermore,
the Coq proof assistant provides not only a great number of tactics
but also a domain-specific language for scripted proof automation,
called $\mathcal{L}$tac. Details about $\mathcal{L}$tac and Coq can be found
in~\cite{Chlipala13,Bertot10,manual_coq}.

@@

\begin{figure}[H]
\begin{enumerate}
  % \item $\mathcal{M}\,;\,w\not\models \bot$ and $\mathcal{M}\,;\,w\models \top$;
  \item $\mathcal{M}\,;\,w\models A$ iff $w \in \mathcal{V}(A)$;
  \item $\mathcal{M}\,;\,w\models \neg\varphi$ iff $\mathcal{M}\,;\,w\not\models\varphi$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \land \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ and $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \lor \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \to \varphi_2$ iff
    $\mathcal{M}\,;\,w\not\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models\Box\varphi$ iff $\forall y \in W. (w R y \to \mathcal{M}\,;\,y \models\varphi)$;
  \item $\mathcal{M}\,;\,w\models\Diamond\varphi$ iff $\exists y \in W. (w R y \land \mathcal{M}\,;\,y \models\varphi)$.
  \end{enumerate}
  \caption{Kripke semantics}
  \label{fig:kripke}
\end{figure}
We say that a formula $\varphi$ is satisfiable in a model $\mathcal{M}$,
denoted by $\mathcal{M}\models \varphi$, if $\forall w \in W. (M\,;w\,\models \varphi)$ holds.
Similarly, $\varphi$ is valid in a frame $\mathcal{F}$, written $\mathcal{F}\models\varphi$, if
it holds on all models $\mathcal{M} = \langle \mathcal{F},\mathcal{V}\rangle$, i.e.,
for any $\mathcal{M}$, we have that $\mathcal{M}\models\varphi$. A formula
is valid, written $\models \varphi$, if it is valid in any frame $\mathcal{F}$. Using the above semantics, we can define the concept of \emph{entailment}: given a set
of formulas $\Gamma$ and a formula $\varphi$, we say that $\Gamma$ entails
$\varphi$, written $\Gamma\models\varphi$, iff $\forall \mathcal{M} . ((\forall \varphi' \in\Gamma .
\mathcal{M} \models \varphi') \to \mathcal{M} \models \varphi)$.
In this definition, if $\Gamma$ entails $\varphi$ it means that $\varphi$ is a global semantic consequence of $\Gamma$, which will be further elaborated on Section~\ref{sec:completeness}.

In this work, we consider Hilbert deductive systems for modal logic.
Essentially, the considered system add axioms for the modalities
to the axioms for classical propositional logic.
As examples of modalities axioms we mention $K$, which express that $\Box$ distributes
over implication, and possibility, which states that $\Diamond$ distributes
over disjunction.
\[
  \begin{array}{ll}
    K & \Box(\varphi_1 \to \varphi_2) \to (\Box \varphi_1 \to \Box \varphi_2)\\
    Possibility & \Diamond (\varphi_1 \lor \varphi_2) \to
                           (\Diamond \varphi_1 \lor \Diamond \varphi_2)\\
  \end{array}
\]
Apart from the axioms, we also need \emph{modus ponens} and necessitation rules.

\begin{figure}[H]
\begin{tabular}{cc}
    \AxiomC{$\Gamma\vdash\varphi_1\to\varphi_2$}
    \AxiomC{$\Gamma\vdash\varphi_1$}
    \RightLabel{MP}
    \BinaryInfC{$\Gamma\vdash\varphi_2$}
    \DisplayProof
  &
    \AxiomC{$\Gamma\vdash\varphi$}
    \RightLabel{Nec}
    \UnaryInfC{$\Gamma\vdash\Box \varphi$}
    \DisplayProof
\end{tabular}
\centering
\end{figure}
The set of axioms and the two previous rules form a sound and complete Hilbert-style
deductive system for modal logics~\cite{blackburn}. Given a formula $\varphi$ and a
set of hypothesis $\Gamma$, the soundness property ensures that
whenever $\varphi$ is deduced from $\Gamma$, then $\Gamma\models \varphi$. On the
other hand, completeness guarantee that if $\Gamma\models\varphi$, then $\varphi$ is
provable from $\Gamma$ using the deductive system. While the soundness argument can
be proved by a immediate inductive argument over derivations of $\Gamma\vdash\varphi$,
the same strategy does not apply to the completeness property: the key step in proving
completeness is to produce a \emph{cannonical model} using maximal consistent sets
of formulas.

@@

Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an instantiation of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
the set of axioms that form a system for some modal logic. In this way, our definition can be
used by several systems alike.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop): theory -> formula -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: formula),
        t f -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: formula),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: formula)
               (d1: deduction A t (Implies f g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: formula)
                (d1: deduction A t f),
         deduction A t ([] f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax} constructor, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : formula -> formula -> axiom
  | ax2  : formula -> formula -> formula  -> axiom
  (* some code omitted *)
  | axK  : formula -> formula -> axiom
  | axPos: formula -> formula -> axiom
  | axT  : formula -> axiom
  | axB  : formula -> axiom
  | axK4 : formula -> axiom
  | axD  : formula -> axiom
  | axK5 : formula -> axiom
  | axGL : formula -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom scheme and
its type stores the subformulas which compose the axiom. The type has
18 constructors: the first 10 encode axioms for propositional logic
and, the last 7 are for different formulas which are taken as axioms by common
systems of modal logic. Axioms are built using the function \mintinline{coq}{instantiate},
which receives an axiom and builds a corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): formula :=
  match a with
  | ax1 p q
     => [! p -> (q -> p) !]
  | ax2 p q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  (* some code omitted *)
  | axK p q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos p q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is useful to use a short notation while still avoiding
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.
%
With such definitions at hand,
the \mintinline{coq}{Ax} constructor uses the \mintinline{coq}{instantiate}
function to create a formula instance from a given system,
represented as a predicate
\mintinline{coq}{A: axiom -> Prop} which parameterizes our deduction type. An
axiom \mintinline{coq}{a} is valid within a system \mintinline{coq}{A} whenever
the type \mintinline{coq}{A a} is inhabited, i.e., it is not empty. Next, we
proceed to
describe how to represent different systems of modal logic.

@@

In this section, we describe our formalization of the completeness result for the
Hilbert proof system with respect to the Kripke semantics.
Before we start,
it is worth mentioning that there are different notions of completeness, and
the literature is not always clear about the difference of those~\cite{Popkorn_1994,modal_lean}.
%-------------------------------------------------------------------------------
The semantic entailment, $\Gamma\models\varphi$, represents a global consequence since, by our definition, we expect that the hypotheses are valid in every world in a model instead of only on those worlds where $\varphi$ hold \cite{de1999global}. Also, the completeness result may be weak, which is only valid without hypotheses, or strong, allowing it to be stated with any set of hypotheses \cite{blackburn}. We follow the definitions as stabilished by Popkorn \cite{Popkorn_1994} and note that we prove a strong completeness result using a global consequence: as this subsumes both weak completeness and a local one (as local consequence implies global consequence \cite{de1999global}), our result allows us to derive other possible definitions of completeness, while still being a weak result enough that it will work for any of the possible sets of axioms \cite{Popkorn_1994}.

@@

\paragraph{Truth as membership in the canonical model}
The key insight to prove the completeness result is to show that truth is membership in the canonical model.
In other words: a formula is true at a world in the canonical model iff it is a member of that world.
This fact is formalized by the truth theorem, which is proved by induction over formula \verb|p|:
\begin{minted}{coq}
Lemma truth: forall p w, (M ' w ||- p) <-> (w p).
\end{minted}
Notation \mintinline{coq}{M ' w ||- p} denotes the valuation of a fórmula at a given world \verb|w| of the
canonical model. Note that a world in the canonical model is an (infinite) consistent set of formulas.
Since we are modelling sets by their characteristic functions (functions of type \mintinline{coq}{formula -> Prop}),
set membership is modelled by function application, \verb|w p|.

@@

In order to state L\"ob's theorem we need a way to encode modal
fixed points. Formally, let $\varphi(X)$ be a formula with only one propositional variable
$X$. The modal fixed point of $\varphi$ is then a formula $\Psi$ such that
$\Psi \leftrightarrow \varphi(\Box \Psi)$.
We represent the existence of modal fixed points for all one-variable formulas by a definition \mintinline{coq}{fixed_point},
which is parameterized by a system of modal logic, \mintinline{coq}{S}, and a
theory, \mintinline{coq}{G}, and encodes the logical equivalence demonstrating
that \mintinline{coq}{p} is a modal fixed point for \mintinline{coq}{f}.
\begin{minted}{coq}
  Definition fixed_point A G: Prop :=
    forall f, exists p,
    (A; G |-- [! (p <-> f ([]p)) !]).
\end{minted}

@@

\begin{minted}{coq}
  Theorem Lob:
    forall A G, subset K4 A /\ fixed_point A G ->
    forall p, (A; G |-- [! []p -> p !]) ->
              (A; G |-- [! p !]).
\end{minted}

@@

\begin{minted}{coq}
  intros A G [I FP] P H1.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; G |-- [! psi -> []psi -> p !]) as H3.
  apply modal_ax5 in H2; auto with modal.
\end{minted}

@@

From $\vdash \Psi \to (\Box \Psi \to p)$ and the necessitation
rule, we can deduce $\vdash \Box (\Psi \to (\Box \Psi \to p))$
using the following tactics:
\begin{minted}{coq}
  assert (A; G |-- [! [](psi -> []psi -> p) !]) as H4.
  apply Nec; auto.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; G |-- [! []psi -> []([]psi -> p) !]) as H5.
  apply modal_axK in H4; auto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; G |-- [! []([]psi -> p) -> [][]psi -> []p !]) as H6.
  eapply Ax with (a := axK ?[X] ?[Y]); auto with modal.
  reflexivity.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; G |-- [! []psi -> [][]psi -> []p !]) as H7.
  eapply modal_compose; eauto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; G |-- [! []psi -> [][]psi !]) as H8.
  apply modal_axK4; auto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; G |-- [! []psi -> []p !]) as H9.
  eapply modal_ax2; eauto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; G |-- [! []psi -> p !]) as H10.
  eapply modal_compose; eauto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; G |-- [! ([]psi -> p) -> psi !]) as H11.
  apply modal_ax6 in H2; auto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; G |-- psi) as H12.
  eapply Mp; try eassumption.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; G |-- [! []psi !]) as H13.
  apply Nec; try eassumption.
\end{minted}

@@

\pagebreak

>>>>

\author{Miguel Nunes}
\ead{miguel.alfredo.nunes@gmail.com}
\address{State University of Campinas}

@@

Such wide range of highly expressive logical languages is the ideal tool to
formalize the many non-trivial properties of modern advanced software systems,
such as those found in cars, airplanes, medical devices and banking applications.
By combining said modal logics with interactive proof assistants,
such as Coq~\cite{manual_coq} and Isabelle/HOL~\cite{manual_isabelle}, both of
which have been applied with success in non-trivial projects like a certified compiler
for C programming language~\cite{Leroy09} and the kernel of an operating
system~\cite{Klein10},
% it is possible to formalize and verify highly complex computer systems.
we argue we may achieve better ways of formally describing and verifying complex programs.

@@

We define next a proof of the previous propositional logical formula
that, in contrast to the previous proof, that was built using tactics, is coded
directly as a function:
\begin{minted}{coq}
Definition example: (A -> B) -> (B -> C) -> A -> C :=
  fun (H: A -> B) (H': B -> C) (HA: A) => H' (H HA).
\end{minted}
However, even for very simple theorems, coding a proof directly
as a Coq term can be a hard task. Because of this, the use of tactics
has become the standard way of proving theorems in Coq. Furthermore,
the Coq proof assistant provides not only a great number of tactics
but also a domain-specific language for scripted proof automation,
called $\mathcal{L}$tac. Details about $\mathcal{L}$tac and Coq can be found
in~\cite{Chlipala13,Bertot10,manual_coq}.

@@

\begin{figure}[H]
\begin{enumerate}
  % \item $\mathcal{M}\,;\,w\not\models \bot$ and $\mathcal{M}\,;\,w\models \top$;
  \item $\mathcal{M}\,;\,w\models A$ iff $w \in \mathcal{V}(A)$;
  \item $\mathcal{M}\,;\,w\models \neg\varphi$ iff $\mathcal{M}\,;\,w\not\models\varphi$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \land \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ and $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \lor \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \to \varphi_2$ iff
    $\mathcal{M}\,;\,w\not\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models\Box\varphi$ iff $\forall y \in W. (w R y \to \mathcal{M}\,;\,y \models\varphi)$;
  \item $\mathcal{M}\,;\,w\models\Diamond\varphi$ iff $\exists y \in W. (w R y \land \mathcal{M}\,;\,y \models\varphi)$.
  \end{enumerate}
  \caption{Kripke semantics}
  \label{fig:kripke}
\end{figure}
We say that a formula $\varphi$ is satisfiable in a model $\mathcal{M}$,
denoted by $\mathcal{M}\models \varphi$, if $\forall w \in W. (M\,;w\,\models \varphi)$ holds.
Similarly, $\varphi$ is valid in a frame $\mathcal{F}$, written $\mathcal{F}\models\varphi$, if
it holds on all models $\mathcal{M} = \langle \mathcal{F},\mathcal{V}\rangle$, i.e.,
for any such $\mathcal{M}$, we have that $\mathcal{M}\models\varphi$. A formula
is valid, written $\models \varphi$, if it is valid in any frame $\mathcal{F}$. Using the above semantics, we can define the concept of \emph{entailment}: given a set
of formulas $\Gamma$ and a formula $\varphi$, we say that $\Gamma$ entails
$\varphi$, written $\Gamma\models\varphi$, iff $\forall \mathcal{M} . ((\forall \varphi' \in\Gamma .
\mathcal{M} \models \varphi') \to \mathcal{M} \models \varphi)$.
In this definition, if $\Gamma$ entails $\varphi$ it means that $\varphi$ is a global semantic consequence of $\Gamma$, which will be further elaborated on Section~\ref{sec:completeness}.

In this work, we consider Hilbert deductive systems for modal logic.
Essentially, the considered system add axioms for the modalities
to the axioms for classical propositional logic.
As examples of modalities axioms we mention $K$, which express that $\Box$ distributes
over implication, and possibility, which states that $\Diamond$ distributes
over disjunction.
\[
  \begin{array}{ll}
    K & \Box(\varphi_1 \to \varphi_2) \to (\Box \varphi_1 \to \Box \varphi_2)\\
    Possibility & \Diamond (\varphi_1 \lor \varphi_2) \to
                           (\Diamond \varphi_1 \lor \Diamond \varphi_2)\\
  \end{array}
\]
Apart from the axioms, we also need \emph{modus ponens} and necessitation rules.
The complete deduction system is as follows:

\begin{figure}[H]
\begin{tabular}{cccc}
    \AxiomC{$\varphi \in \Gamma$}
    \RightLabel{Prem}
    \UnaryInfC{$\Gamma\vdash \varphi$}
    \DisplayProof
  &
    \AxiomC{$\varphi\:\textit{is an axiom}$}
    \RightLabel{Ax}
    \UnaryInfC{$\Gamma \vdash \varphi$}
    \DisplayProof
  &
    \AxiomC{$\Gamma\vdash\varphi_1\to\varphi_2$}
    \AxiomC{$\Gamma\vdash\varphi_1$}
    \RightLabel{MP}
    \BinaryInfC{$\Gamma\vdash\varphi_2$}
    \DisplayProof
  &
    \AxiomC{$\vdash\varphi$}
    \RightLabel{Nec}
    \UnaryInfC{$\Gamma\vdash\Box \varphi$}
    \DisplayProof
\end{tabular}
\centering
\end{figure}
\noindent In the SBLP 2022 version of this work, we use a different deductive system in which
the necessitation rule was as follows:
\[
    \AxiomC{$\Gamma\vdash\varphi$}
    \RightLabel{Nec}
    \UnaryInfC{$\Gamma\vdash\Box \varphi$}
    \DisplayProof
\]
However, this system does not admit the deduction theorem. Since we aim for a
completeness result and it implies the deduction theorem, the system which we used in
the previous version can't be complete and hence requires this changes.
For a clarification on the change and the role of the deduction theorem in modal logic,
we refer the reader to~\cite{HakliN12}.
The set of axioms and the two last rules form a sound and complete Hilbert-style
deductive system for modal logics~\cite{blackburn}. Given a formula $\varphi$ and a
set of hypothesis $\Gamma$, the soundness property ensures that
whenever $\varphi$ is deduced from $\Gamma$, then $\Gamma\models \varphi$. On the
other hand, completeness guarantee that if $\Gamma\models\varphi$, then $\varphi$ is
provable from $\Gamma$ using the deductive system. While the soundness argument can
be proved by a immediate inductive argument over derivations of $\Gamma\vdash\varphi$,
the same strategy does not apply to the completeness property: the key step in proving
completeness is to produce a \emph{cannonical model} using maximal consistent sets
of formulas.

@@

Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an instantiation of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
the set of axioms that form a system for some modal logic. In this way, our definition can be
used by several systems alike.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop): theory -> formula -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: formula),
        t f -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: formula),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: formula)
               (d1: deduction A t (Implies f g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: formula)
                (d1: deduction A Empty i),
         deduction A t ([] f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax} constructor, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : formula -> formula -> axiom
  | ax2  : formula -> formula -> formula  -> axiom
  (* some code omitted *)
  | axK  : formula -> formula -> axiom
  | axPos: formula -> formula -> axiom
  | axT  : formula -> axiom
  | axB  : formula -> axiom
  | axK4 : formula -> axiom
  | axD  : formula -> axiom
  | axK5 : formula -> axiom
  | axGL : formula -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom scheme and
its type stores the subformulas which compose the axiom. The type has
18 constructors: the first 10 encode axioms for propositional logic
and, the last 7 are for different formulas which are taken as axioms by common
systems of modal logic. Axioms are built using the function \mintinline{coq}{instantiate},
which receives an axiom and builds a corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): formula :=
  match a with
  | ax1 p q
     => [! p -> (q -> p) !]
  | ax2 p q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  (* some code omitted *)
  | axK p q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos p q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is useful to use a short notation while still avoiding
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.
%
With such definitions at hand,
the \mintinline{coq}{Ax} constructor uses the \mintinline{coq}{instantiate}
function to create a formula instance from a given system,
represented as a predicate
\mintinline{coq}{A: axiom -> Prop} which parameterizes our deduction type. An
axiom \mintinline{coq}{a} is valid within a system \mintinline{coq}{A} whenever
the type \mintinline{coq}{A a} is inhabited, i.e., it is not empty. Next, we
proceed to
describe how to represent different systems of modal logic.

@@

In this section, we describe our formalization of the completeness result for the
Hilbert proof system with respect to the Kripke semantics.
Before we start,
it is worth mentioning that there are different notions of completeness, and
the literature is not always clear about the difference of those~\cite{Popkorn_1994,modal_lean}.
%-------------------------------------------------------------------------------
The semantic entailment, $\Gamma\models\varphi$, represents a global consequence since, by our definition,
we expect that the hypotheses are valid in every world in a model instead
of having $\varphi$ hold only on those worlds where the hypotheses are valid \cite{de1999global}.
Also, the completeness result may be
weak, which is only valid without hypotheses, or strong, allowing it to be stated with any set of
hypotheses \cite{blackburn}. We follow the definitions as stabilished by Popkorn \cite{Popkorn_1994}
and note that we prove a strong completeness result using a global consequence: as this subsumes
both weak completeness and a local one (as local consequence implies global consequence
\cite{de1999global}), our result allows us to derive other possible definitions of completeness,
while still being a weak result enough that it will work for any of the possible sets of
axioms \cite{Popkorn_1994}.

@@

\paragraph{Truth as membership in the canonical model}
The key insight to prove the completeness result is to show that truth is membership in the canonical model.
In other words: a formula is true at a world in the canonical model iff it is a member of that world.
This fact is formalized by the truth theorem, which is proved by induction over formula \verb|p|:
\begin{minted}{coq}
Lemma truth: forall p w, (M ' w ||- p) <-> (w p).
\end{minted}
Notation \mintinline{coq}{M ' w ||- p} denotes the valuation of a formula at a given world \verb|w| of the
canonical model. Note that a world in the canonical model is an (infinite) consistent set of formulas.
Since we are modelling sets by their characteristic functions (functions of type \mintinline{coq}{formula -> Prop}),
set membership is modelled by function application, \verb|w p|.

@@

In order to state L\"ob's theorem we need a way to encode modal
fixed points. Formally, let $\varphi(X)$ be a formula with only one propositional variable
$X$. The modal fixed point of $\varphi$ is then a formula $\Psi$ such that
$\Psi \leftrightarrow \varphi(\Box \Psi)$.
We represent the existence of modal fixed points for all one-variable formulas by a definition \mintinline{coq}{fixed_point},
which is parameterized by a system of modal logic, \mintinline{coq}{A}, and a
theory, \mintinline{coq}{G}, and encodes the logical equivalence demonstrating
that \mintinline{coq}{p} is a modal fixed point for \mintinline{coq}{f}.
\begin{minted}{coq}
  Definition fixed_point A G: Prop :=
    forall f, exists p,
    (A; G |-- [! (p <-> f ([]p)) !]).
\end{minted}

@@

\begin{minted}{coq}
  Theorem Lob:
    forall A, subset K4 A /\ fixed_point A Empty ->
    forall p, (A; Empty |-- [! []p -> p !]) ->
              (A; Empty |-- [! p !]).
\end{minted}

@@

\begin{minted}{coq}
  intros A [I FP] P H1.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; Empty |-- [! psi -> []psi -> p !]) as H3.
  apply modal_ax5 in H2; auto with modal.
\end{minted}

@@

From $\vdash \Psi \to (\Box \Psi \to p)$ and the necessitation
rule, we can deduce $\vdash \Box (\Psi \to (\Box \Psi \to p))$
using the following tactics:
\begin{minted}{coq}
  assert (A; Empty |-- [! [](psi -> []psi -> p) !]) as H4.
  apply Nec; auto.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; Empty |-- [! []psi -> []([]psi -> p) !]) as H5.
  apply modal_axK in H4; auto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; Empty |-- [! []([]psi -> p) -> [][]psi -> []p !]) as H6.
  eapply Ax with (a := axK ?[X] ?[Y]); auto with modal.
  reflexivity.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; Empty |-- [! []psi -> [][]psi -> []p !]) as H7.
  eapply modal_compose; eauto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; Empty |-- [! []psi -> [][]psi !]) as H8.
  apply modal_axK4; auto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; Empty |-- [! []psi -> []p !]) as H9.
  eapply modal_ax2; eauto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; Empty |-- [! []psi -> p !]) as H10.
  eapply modal_compose; eauto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; Empty |-- [! ([]psi -> p) -> psi !]) as H11.
  apply modal_ax6 in H2; auto with modal.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; Empty |-- psi) as H12.
  eapply Mp; try eassumption.
\end{minted}

@@

\begin{minted}{coq}
  assert (A; Empty |-- [! []psi !]) as H13.
  apply Nec; try eassumption.
\end{minted}

@@
